Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-18 Manuscript/manuscript.Rmd
manuscript edits

**Background.** Advances in Hi-C sequencing technologies have revealed the hierarchical structure of genome organization. Among this organization exists regions in the genome that are highly self-interacting, termed topologically associated domains (TADs). The locations of TADs throughout the linear genome has been studied extensively. However, the set of functional genomic annotations that are associated with the formation of TADs has not yet been systematically evaluated. By identifying the annotations that are enriched near regions that demarcate TADs, we can gain a better understanding of the mechanisms that lead to the 3D structure of the genome.

Current methods rely on the use of classification algorithms to predict influential annotations. However, key characteristics of domain data are typically ignored. Most importantly, the sparsity of TAD boundaries throughout the linear genome leads to heavily imbalanced classes that distinguish boundary vs. non-boundary regions. In addition, determining the most informative relationship between TAD boundaries and genomic annotations remains an open problem. 
**Methods.** To address this, we proposed a novel ensemble framework in order to systematically compare the performances of random forest classifiers implemented on combinations of different Hi-C contact matrix resolutions, feature engineering strategies, and re-sampling techniques.

**Results.** Using the GM12878 and K562 cell lines, we show that models built on SMOTE (Synthetic Minority Over-sampling TEchnique) re-sampled data out performed models built simply on random under- and over-sampled data, across all resolutions. Additionally, our results indicated that distances between annotations and TAD boundaries were more informative than other predictor-types used to build the feature space of models. We also observed that performances tended to degrade as data resolution decreased. Lastly, the most predictive annotations in the best performing models were found to be consistent across both cell line and resolution. They included well-known marks of TAD boundaries such as CTCF, SMC3, RAD21, and ZNF143, as well as lesser known transcription factors including P300 and NFKB.

**Conclusions.** The experiments derived from our ensemble framework provide a useful guide toward the exploration of chromatin organization, while highlighting the impact of both class imbalance and feature engineering in TAD boundary prediction. Our framework provides insight into the best set of parameters necessary for optimal performance.

@@

After determining the optimal set of data level characteristics necessary for predicting TAD boundaries, we then aimed to identify the set of functional genomic elements that were most influential in predicting said boundaries.

In order to assess the predictive capabilities of specific types of annotations, we further compared the performances of annotation-specific models. Using the optimal set of characteristics obtained above (SMOTE re-sampling with distance-type predictors), random forest classifiers built on sets of cell line specific histone modifications, chromatin state segmentations, and transcription factor binding sites (TFBS) were compared to a model that included all three types together. For GM12878, models built simply on TFBS were found to perform consistently just as well as models that included all 3 sets of annotations, across resolution (Figure 6). Similar results were found when evaluating the K562 cell line (supplementary Figure 11). For both resolutions, models built on only histone modifications were found to have the lowest MCC, across resolutions, followed by models built on chromatin state segmentations.

@@

As an external method of validation, we also investigated the predictive importance of the annotations that our top performing models were built on. We found that specific TFBS, including SMC3ab, ZNF143, CTCF, and RAD21, were given consistently high predictive importance in the models built on the GM12878 cell line, across all resolutions (Figure 8). These are known factors associated with the hypothesized loop extrusion model [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self] and were found to yield consistently high predictive importance on the K562 cell line as well, with similar profiles across resolutions (Supplementary Figure 15). [???mention additional tfbs when full results ready]

>>>>

**Background.** Topologically Associated Domains (TADs) represent fundamental building blocks of the three-dimensional (3D) genome organization and regulation. Several epigenetic marks, such as CTCF, are highly enriched at TAD boundaries, suggesting a strong epigenetic component of TAD boundary formation. Several methods have been developed to leverage the growing amount of (epi)genomic annotation data for predicting TAD boundaries; however, they overlooked key characteristics of the data, such as spatial associations between a genomic annotation and a boundary. Furthermore, the number of regions containing TAD boundaries is much smaller than the rest of the genome, creating a class imbalance problem.

**Methods.** To investigate the problem of spatial genomic relationships and class imbalance problem, we developed an ensemble framework for optimizing prediction of TAD boundaries. We utilized several resampling techniques (random over- and undersampling, Synthetic Minority Over-sampling TEchnique (SMOTE)) coupled with various metrics for classifier performance evaluation (Area Under Receiver Operating Characteristics (AUCROC) curve, Precision-Recall curve (AUPRC), F1, Matthew Correlation Coefficient (MCC), among others) to select the best predictive model in class imbalance settings.

**Results.** Using different feature engineering and resampling strategies, coupled with elastic net-based feature selection, we systematically compared the performance of models predicting the location of TAD boundaries (Gm12878 and K562 Hi-C data at 10kb to 100kb resolutions). Precision-Recall curves were found to be best suited for evaluating model performance in class imbalance settings. Random undersampling and SMOTE-resampled data performed best for optimal parameter selection. Transcription Factor Binding Sites (TFBSs) were best type of genomic annotations for predicting TAD boundaries than histone modifications and chromatin states; we identified well-known CTCF, SMC3, RAD21, and several novel TFBSs as best predictors of TAD boundaries. Importantly, spatial relationships (distance between TAD boundaries and genomic annotations) were best predictors of TAD boundaries as compared with commonly used overlap counts or percent overlap measures.

**Conclusions.** Our results outline strategies for predictive modeling of 3D genomic domains using 1D genome annotation data. The accurate modeling of TAD boundaries will improve our understanding of how epigenetics shapes the 3D structure of the genome.

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites), across resolution. We found that for random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the cell-line specific transcription factor binding sites, and was consistent across resolutions (Figure 5; Supplementary Figure 4).

We further assessed the predictive capabilities of the specific sets of annotations by comparing the performances of annotation-specific random forest models implemented on the optimal set of characteristics. We found that models built simply on cell line specific TFBS out performed other cell line and annotation-specific models at all resolutions (Figure 6; Supplementary Figure 5). Additionally, when compared to the optimal model identified from the ensemble framework, one that included all 3 sets of annotations, the TFBS-specific models performed similarly at the 10 kb and 25 kb resolutions. Interestingly though, at the 50 kb and 100 kb resolutions, the TFBS-specific models out performed models built on the combined sets of annotations, indicating that, at lower resolution, the inclusion of other types of annotations contributed to additional noise rather than signal.

@@

We then sought to determine which of the transcription factor binding sites were most influential in predicting TAD boundaries. We found that the cell line specific binding sites associated with SMC3, CTCF, RAD21, ZNF143, and YY1 were given consistently high predictive importance, across all resolutions (Figure 7; Supplementary Figure 6). The SMC3, CTCF, and RAD21 binding sites are known factors associated with the cohesin protein complex, which is hypothesized to be what creates loops in the DNA  [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likwise, ZNF143 and YY1 are known structural regulators of enhancer-promoter interactions that also contribute to looping of the linear genome [@weintraub2017yy1; @gonzalez2017znf143]. With the identification of these factors known to be enriched at TAD boundaries, we are able to confirm the validity of our model building process. Additionally, we were also able to identify some lesser known cell line specific transcription factors as predictive of TAD boundaries including the MYC-associated zing-finger protein (MAZ), jun-D proto-oncogene (JUND), and ELF1. These transcription factor binding sites tended to be given higher predictive importance at lower resolution. In fact, at lower resolutions, the predictive importance was found to be much more evenly dispersed across the top transcription factor binding sites, than for higher resolutions. This is again suggestive of the fact that the spatial relationship between functional annotations, specifically transcription factor binding sites, and TAD boundaries changes at different Hi-C data resolutions.
